Synthesizing Quotient Lenses

### 2018-03-13 papers/QuotientLenses/main.tex
minor tweak
Checkpoint some tweaking and comments, mostly toward the beginning

\lstset{ language=Caml, basicstyle=\linespread{0.8}\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkblue}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=false,
commentstyle=\itshape\rmfamily,
emph={synth,project,perm,squash,normalize,using,ins,del,lens},
emphstyle=\upshape\sffamily\color{dkblue}, columns=fullflexible,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
mathescape, }

@@

This paper simplifies the task of programming quotient lenses with
three connected innovations.
First, we introduce {\em quotient regular expressions} (QREs), to
easily mark aspects of data formats that are
inessential.  A QRE is an augmented regular expression that
simultaneously defines a regular language and an equivalence
relation on this language.  Variations within the same
equivalence class are deemed inessential. 
Next, we introduce {\em QRE lenses}---lenses mapping
between QREs.
Our key technical result is a proof that every QRE lens can be
transformed into a QRE lens that has the form $c_1 \cdot \ell \cdot
c_2$, where $c_1$ and $c_2$ are {\em canonizers} 
and $\ell$ is a bijective transformation.  Intuitively,
the canonizers normalize data by picking a
canonical representative of the relevant equivalence relations. 
This theorem means we can push all normalization to the outer edges of a
bidirectional transformation, leaving in the middle a pure bijective
transformation.  
Finally,
we leverage this normalization theorem to {\em synthesize} QRE lenses from a pair
of QREs and example input-output pairs, reusing earlier work on
synthesizing plain bijective lenses.
We have implemented QREs and QRE lens synthesis 
in the bidirectional programming language Boomerang.
We evaluate the effectiveness of our approach by using it to
synthesize QRE lenses between various real-world data formats in
the Optician benchmark suite.

@@

\section{Introduction}
Often, the same data is stored in more than one format. For example,
bibliographic data may be stored in \bibtex{} or EndNote formats,
spreadsheet data can be represented as either CSV and TSV files, and web
APIs can return data as either JSON or XML objects.  When changes are
made to data in one format, data in the other must be updated so the
two representations continue to store the same information.
One way to synchronize such data is to use \emph{lenses}~\cite{Lenses}, which
are programs that simultaneously define pairs of functions for translating
the data from the source format to the target formant (the \emph{get}
direction) and back again (the \emph{put} direction). Lenses are usually
defined using domain-specific bidirectional languages that 
ensure the specified lenses satisfy various \emph{lens laws} that
guarantee how the \emph{put} and \emph{get} functions interact.

\emph{Bijective lenses} are amongst the most restrictive. 
A bijective lens $\ell : S \Leftrightarrow T$ specifies a bijection from
a source set $S$ to a target set $T$:
\begin{equation}\label{bijectivelenslaws}
\ell.\get \; (\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s)
= s
\end{equation}
Since these lenses demand that the \emph{get} and \emph{put} functions form a bijection, 
no information is lost during translation from one data format to another. 
While this very strong guarantee can be useful, more often
some of the information in one or both of the data formats is
superfluous. For instance, the number of white space characters
between data items, the capitalization of various strings, the
sequence of fields in a record, or the order of items in a list are
often irrelevant.  Many
data formats have their own ad hoc rules as well. In \bibtex{}, for example,
an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---
the specific choice does not matter.

{\em Quotient lenses}~\cite{quotientlenses} loosen the restrictions of
bijective string lenses by including equivalence
relations on the source and target formats so that the round-tripping lens laws
hold modulo these equivalences rather than modulo equality. Quotient lenses thus
make it possible to synchronize many more data formats than would otherwise be
possible by ignoring details deemed inessential.

@@

Our solution to this shortcoming is to introduce {\em Quotient Regular
Expressions} (QREs), which enable programmers to simultaneously specify a
regular expression and a set of canonical representatives for matching
strings.  Given an QRE, we can automatically infer the corresponding
canonization functions.  As an example, consider the following QRE for writing
author names:
%
\begin{lstlisting}
let wsp_sp = project wsp $\to$ " "
let comma_name = first_name . "," . wsp_sp . last_name
\end{lstlisting}
%
In this example, \lstinline{wsp} is an existing regular expression for a
nonempty sequence of whitespace characters, and \lstinline{first_name} and
\lstinline{last_name} are existing regular expressions for first and last names.
The QRE \lstinline{wsp_sp} is a QRE with the same underlying language as
\lstinline{wsp}, but with a single canonical representative: \lstinline{" "}.
This is then used to build up \lstinline{comma_name}, a QRE with an
underlying language of two names, separated by a command and whitespace, where
the names with a single space between them are canonical.

@@

\section{Background: Bijective String Languages}
\label{sec:background}
\ksf{We are inconsistent in how we typeset get/put.  Sometimes we use
codefont, sometimes we use italics of various flavors, and sometimes
we don't do anything distinctive.  It would be good to be consistent!}
Before describing QREs and QRE lenses, we briefly review bijective
string lenses. Consider a bidirectional transformation that converts
between \bibtex{} citation records such as
\begin{verbatim}
  @Book {Lovelace,
  Author = "Ada Lovelace",
  Title = {Notes}
}
\end{verbatim}
\noindent
and equivalent EndNote records like the following
\begin{verbatim}
  %0 Book
  %F Lovelace
  %A Ada Lovelace
  %T Notes
\end{verbatim}
\noindent
Boomerang's bijective string lenses are designed to define bidirectional
transformations such as this one, where the data formats can be matched in a
one-to-one manner, in this case by matching the label, author and title fields.
Boomerang encourages a compositional approach in which programmers
define simple lenses and then compose them using a variety of
combinators.

@@

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions, which are omitted for brevity.  
\ifdraft
\bcp{And a readability suggestion:
  define a named constant for a quote character.}
\fi}
\label{fig:example-lens}
\end{figure}

@@

\begin{lstlisting}
let wsp_sp = project wsp $\to$ " "
\end{lstlisting}
\iflastminute
\bcp{Can we display blanks as a distinct ``visible blank'' symbol
(throughout)?  That will make examples easier to read.}\bcp{I tried turning
on the showstringspaces option at the top, but that doesn't do everything we
want--it only works inside quotes.  But the latex ``textvisiblespace'' command
produces a single one, and the lstlisting ``escapechar'' option will allow
us to use it inside listings.  But we should save this for the final
tweaking stage.}
% Here is an example: if you set escapechar=\%, you could use it inside the
% listings as follows: \begin{lstlisting} t %\leftarrow% 0 \end{lstlisting}
\fi

@@

For instance, assume that $f(s)=$ `` '' for all whitespace strings $s$. Then the
project QRE \lstinline{wsp_sp} defined above can
be expressed using the normalize combinator as
\lstinline{normalize (wsp," ", f)}.

Figure~\ref{fig:example-qre} gives the complete QRE specification for the
\bibtex{}.
\begin{figure}
\begin{lstlisting}
let wsp = [ \n\t\r]+
let wsp_sp = project wsp $\to$ " "
let last_name = [A-Z][a-z]
let first_name = [A-Z][a-z]

@@

\section{Quotient Regular Expressions}
\label{QRE}
A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section describes in further detail the set of QRE combinators that we described
in \cref{sec:example}.\saz{Why does the ``cref'' macro not capitalize Section?}

@@

Semantically, the \kw{normalize} QRE is universal---each of the other
combinators $Q$ is equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some
surjective, idempotent function
$f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$.   However,
although \kw{normalize} is powerful, verifying that a canonization function $f$
is surjective and idempotent is in general undecidable.  This means that a
programmer wishing to use \kw{normalize} must discharge strong proof
obligations, which is cumbersome in practice.\footnote{In our implementation we allow a programmer to use \kw{normalize} at their
own risk without checking these side conditions as an ``escape hatch'' for the
case when other QRE combinators are insufficient.}

@@

\begin{figure}[t]
\centering
\[
\begin{array}{l@{}l@{\ \ }l}
w \; \equiv_{\id(R)} \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' && \text{for all }w, w'\\
w \; \equiv_{\squash{Q_1}{Q_2}{f}} \; w' &\iff& f(w) \equiv_{Q_2} w'
\text{ or } w \equiv_{Q_2} w' \\
w \; \equiv_{\normalize{R}{R'}{f}} \; w' &\iff&
f(w)=f(w') \text{ or }w = w'\\
w \; \equiv_{Q_1 \; ; \; Q_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(Q_2)) \text{ such that } w \; \equiv_{Q_1} \; k, \; w' \;
\equiv_{Q_1} \; k', \text{ and } k \; \equiv_{Q_2} \; k'\\
w \; \equiv_{Q_1 \cdot Q_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{Q_1}
\; {r'}_1, \; r_2 \; \equiv{Q_n} \; {r'}_2\\
w \; \equiv_{Q_1 \sep Q_2} \; w' &\iff& w \; \equiv_{Q_1} \; w'
\text{ or } \; w \; \equiv_{Q_2} \; w'\\
w \; \equiv_{Q^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{Q} \; {r'}_i
\\
w \; \equiv_{\perm{Q_1, \ldots, Q_n}{Q}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{Q} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Ambiguity and Well Formed QREs}
\label{subsec:well-formed-qres}
An issue that we address before presenting the full set of inference
rules for QREs is the issue of {\em ambiguity}, which appears in the
premises of the inference rules for the regular combinators. If $Q_1$ and $Q_2$
are QREs, then when applying the regular combinators to QREs, we require that
if a string $s$ matches any of the regular expressions $W(Q_1) \cdot W(Q_2)$,
\quad $W(Q_1) \sep W(Q_2)$, \quad $W(Q_1)^*$, \quad $K(Q_1) \cdot K(Q_2)$,
\quad $K(Q_1) \sep K(Q_2)$, \quad $K(Q_1)^*$, then $s$ matches that regular
expression in only one way.

>>>>

\lstset{ language=Caml, basicstyle=\linespread{0.8}\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkblue}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=true,
commentstyle=\itshape\rmfamily,
emph={synth,project,perm,squash,normalize,using,ins,del,lens,let},
emphstyle=\upshape\sffamily\color{dkred}, 
columns=fullflexible,
escapechar=\#,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
mathescape, }

@@

We offer three interconnected innovations to simplify the task of programming
quotient lenses.  
First, we introduce {\em quotient regular expressions} (QREs), to
easily mark aspects of data formats that are
inessential.  A QRE is an augmented regular expression that
simultaneously defines a regular language and an equivalence
relation on this language.  Variations within the same
equivalence class are deemed inessential. 
Next, we introduce {\em QRE lenses}, i.e., lenses mapping
between QREs.
Our key technical result is a proof that every QRE lens can be
transformed into one of the form $c_1 ; \ell ;
c_2^{-1}$, where $c_1$ and $c_2^{-1}$ are {\em canonizers}, $\ell$ is a bijective
transformation, and $;$ is composition\iflastminute\bcp{$c_2^{-1}$ is ugly, but I think
  we need the superscript for accuracy.  Perhaps calling them $c$ and $d$
  instead of $c_1$ and $c_2$ would be better?}\fi.  Intuitively,
the canonizers normalize data by picking a
canonical representative of the relevant equivalence relations. 
This theorem means we can push all normalization to the outer edges of a
bidirectional transformation, leaving in the middle a pure bijective
transformation.  
Finally,
we leverage this normalization theorem to {\em synthesize} QRE lenses from a pair
of QREs and example input-output pairs, reusing earlier work on
synthesizing plain bijective lenses.
We have implemented QREs and QRE lens synthesis 
in the bidirectional programming language Boomerang.
We evaluate the effectiveness of our approach by using it to
synthesize QRE lenses between various real-world data formats in
the Optician benchmark suite.

@@

\section{Introduction}
Often, the same data is stored in more than one format. For example,
bibliographic data may be stored in \bibtex{} or EndNote formats,
spreadsheet data can be represented as either CSV and TSV files, and web
APIs can return data as either JSON or XML objects.  When changes are
made one version, the data in the other must be updated so the
two continue to represent the same information.
One way to synchronize such data is to use \emph{lenses}~\cite{Lenses}, 
programs that simultaneously define pairs of functions for translating
the data from the source format to the target formant (the \emph{get}
direction) and back again (the \emph{put} direction). Lenses can be
written in one of several domain-specific bidirectional languages that 
guarantee various \emph{lens laws} constraining how the \emph{put} and
\emph{get} functions must fit together.

\emph{Bijective lenses} are amongst the most restrictive. 
A bijective lens $\ell : S \Leftrightarrow T$ specifies a bijection from
a source set $S$ to a target set $T$:
\begin{equation}\label{bijectivelenslaws}
\ell.\get \; (\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s)
= s
\end{equation}
Since the \emph{get} and \emph{put} functions form a bijection, 
no information is lost during translation from one data format to another. 
While this very strong guarantee can sometimes be useful, it is more common for 
some of the details in one or both of the data formats to be
superfluous. For instance, the number of white space characters
between data items, the capitalization of various strings, the
sequence of fields in a record, or the order of items in a list are
often irrelevant.  Also, many
data formats introduce their own ad hoc conventions about formatting details
that are taken to be inessential. In \bibtex{}, for example,
an author can be specified either as 
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''.

{\em Quotient lenses}~\cite{quotientlenses} loosen the strong restrictions of
bijective string lenses by including equivalence
relations on the source and target formats and requiring that the
round-tripping lens laws 
hold  modulo these equivalences, rather than ``on the nose.'' Quotient lenses 
make it possible to synchronize many more data formats than is possible with
plain bijective lenses\bcp{Do we justify this claim later?}.

@@

Our solution to this shortcoming is to introduce {\em Quotient Regular
Expressions} (QREs), which enable programmers to simultaneously specify a
regular expression and a set of canonical representatives for matching
strings.  Given an QRE, we can automatically infer the corresponding
canonization functions.  As an example, consider the following QRE for writing
author names:
\iflastminute
\bcp{I'd love to use a visible space symbol for important space characters
  -- this would make the code examples very much clearer!  But it will
  require a bit of hacking: We can use showstringspaces=false}
\fi
%
\begin{lstlisting}
let wsp_sp = project wsp $\to$ " "
let comma_name = first_name . "," . wsp_sp . last_name
\end{lstlisting}
%
In this example, \lstinline{wsp} is an existing regular expression for a
nonempty sequence of whitespace characters, and \lstinline{first_name} and
\lstinline{last_name} are existing regular expressions for first and last names.
The QRE \lstinline{wsp_sp} is a QRE with the same underlying language as
\lstinline{wsp}, but with a single canonical representative: \lstinline{" "}.
This is then used to build up \lstinline{comma_name}, a QRE with an
underlying language of two names, separated by a command and whitespace, where
the names with a single space between them are canonical.

@@

\section{Background: Bijective String Languages}
\label{sec:background}
\iflastminute
\ksf{We are inconsistent in how we typeset get/put.  Sometimes we use
codefont, sometimes we use italics of various flavors, and sometimes
we don't do anything distinctive.  It would be good to be consistent!}
\fi
Before describing QREs and QRE lenses, we briefly review bijective
string lenses. Consider a bidirectional transformation that converts
between \bibtex{} citation records such as
\begin{verbatim}
  @Book {Lovelace,
  Author = "Ada Lovelace",
  Title = {Notes}
}
\end{verbatim}
\noindent
and equivalent EndNote records like the following
\begin{verbatim}
  %0 Book
  %F Lovelace
  %A Ada Lovelace
  %T Notes
\end{verbatim}
\noindent
Boomerang's bijective string lenses are designed to define bidirectional
transformations such as this one, where the data formats can be matched in a
one-to-one manner, in this case by matching the label, author and title fields.
Boomerang encourages a compositional approach in which programmers
define simple lenses and then compose them using a variety of
combinators.

@@

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions, which are omitted for brevity.  
\iflastminute
\bcp{a possible readability suggestion: define a named constant for a quote
  character.  (Not sure if this makes sense, or if it improves things...)}
\fi}
\label{fig:example-lens}
\end{figure}

@@

\begin{lstlisting}
let wsp_sp = project wsp $\to$ " "
\end{lstlisting}
\iflastminute
\bcp{Can we display blanks as a distinct ``visible blank'' symbol
(throughout)?  That will make examples easier to read.}\bcp{I tried turning
on the showstringspaces option at the top, but that doesn't do everything we
want--it only works inside quotes.  But the latex ``textvisiblespace'' command
produces a single one, and the lstlisting ``escapechar'' option will allow
us to use it inside listings.  But we should save this for the final
tweaking stage.}\bcp{Done, I think.}
% Here is an example: if you set escapechar=\%, you could use it inside the
% listings as follows: \begin{lstlisting} t %\leftarrow% 0 \end{lstlisting}
\fi

@@

For instance, assume that $f(s)=$ ``\textvisiblespace'' (a space
character) for all whitespace strings $s$. Then the 
project QRE \lstinline{wsp_sp} defined above can
be expressed using the normalize combinator as
\lstinline{normalize (wsp," ", f)}.

Figure~\ref{fig:example-qre} gives the complete QRE specification for the
\bibtex{}.
\begin{figure}
\begin{lstlisting}
let wsp = [#\,\textvisiblespace\,#\n\t\r]+
let wsp_sp = project wsp $\to$ " "
let last_name = [A-Z][a-z]
let first_name = [A-Z][a-z]

@@

\section{Quotient Regular Expressions}
\label{QRE}
A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section describes in further detail the set of QRE combinators that we described
in \cref{sec:example}.
\iflastminute 
\saz{Why does the ``cref'' macro not capitalize Section?}
\bcp{Dunno.  Because SIGPLAN does not want it to??  But maybe we should be
  using autoref instead anyway...}
\fi

@@

Semantically, the \kw{normalize} QRE is universal---each of the other
combinators $Q$ is equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some
surjective, idempotent function
$f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$.   However,
although \kw{normalize} is powerful, verifying that a canonization function $f$
is surjective and idempotent is in general undecidable.  This means that a
programmer wishing to use \kw{normalize} must discharge strong proof
obligations, which is cumbersome in practice.\footnote{In our implementation we allow a programmer to use \kw{normalize} at their
own risk without checking these side conditions as an ``escape hatch'' for the
case when other QRE combinators are insufficient.\bcp{Do we use this in any
  of the programs we've written?}}

@@

\begin{figure}[t]
\centering
\[
\begin{array}{l@{}l@{\ \ }l}
w \; \equiv_{\normalize{R}{R'}{f}} \; w' &\iff&
f(w)=f(w') \\
w \; \equiv_{\id(R)} \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' && \text{for all }w, w'\\
w \; \equiv_{\squash{Q_1}{Q_2}{f}} \; w' &\iff& f(w) \equiv_{Q_2} w'
\text{ or } w \equiv_{Q_2} w' \\
w \; \equiv_{Q_1 \; ; \; Q_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(Q_2)) \text{ such that } w \; \equiv_{Q_1} \; k, \; w' \;
\equiv_{Q_1} \; k', \text{ and } k \; \equiv_{Q_2} \; k'\\
w \; \equiv_{Q_1 \cdot Q_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{Q_1}
\; {r'}_1, \; r_2 \; \equiv{Q_n} \; {r'}_2\\
w \; \equiv_{Q_1 \sep Q_2} \; w' &\iff& w \; \equiv_{Q_1} \; w'
\text{ or } \; w \; \equiv_{Q_2} \; w'\\
w \; \equiv_{Q^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{Q} \; {r'}_i
\\
w \; \equiv_{\perm{Q_1, \ldots, Q_n}{Q}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{Q} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Ambiguity and Well Formed QREs}
\label{subsec:well-formed-qres}
An issue that we address before presenting the full set of inference
rules for QREs is the issue of {\em ambiguity}, which appears in the
premises of the inference rules for the regular combinators. If $Q_1$ and $Q_2$
are QREs, then when applying the regular combinators to QREs, we require that
if a string $s$ matches any of the regular expressions $W(Q_1) \cdot W(Q_2)$,
\quad $W(Q_1) \sep W(Q_2)$, \quad $W(Q_1)^*$, \quad $K(Q_1) \cdot K(Q_2)$,
\quad $K(Q_1) \sep K(Q_2)$, \quad $K(Q_1)^*$, then $s$ matches that regular
expression in only one way.
